\documentclass[sigconf,nonacm]{acmart}

\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}

\title{Learning Optimization Skills for GPU Kernels: \\
A Knowledge--Environment--Objective Framework}

\author{Haibin Lai}
\affiliation{%
  \institution{Zhejiang University / Microsoft}
}
\email{haibinlai@zju.edu.cn}

\begin{document}

\begin{abstract}
Optimizing GPU kernels for high-performance computing (HPC) workloads remains a labor-intensive process that relies on expert knowledge accumulated over years of practice. Existing approaches either encode this knowledge as static rules or resort to brute-force search, failing to capture the reusable structure inherent in optimization strategies. We observe that a small set of \emph{optimization skills}---transferable patterns such as tiled GEMM with online reduction, temporal blocking, and SMEM-cached streaming---can cover a large fraction of the optimization space across diverse computational patterns. Based on this insight, we propose a framework that automatically \emph{discovers} optimization skills from profiling-guided optimization trajectories, \emph{abstracts} them into reusable (condition, pattern, transformation) triples, and \emph{transfers} them to new kernels. We formalize this through a Knowledge--Environment--Objective (KEO) cognitive architecture where an LLM-driven agent interacts with hardware profilers (NCU), domain knowledge (papers/libraries), and multi-scale benchmarks. Experiments on 10~operators covering all 7 Berkeley Dwarfs on an NVIDIA A100 GPU show that 5 automatically discoverable skills achieve 1.2--6.0$\times$ speedup over materialized baselines, and that skills learned from one operator transfer to unseen operators with minimal adaptation.
\end{abstract}

\keywords{GPU optimization, kernel compilation, optimization skills, LLM agents, hardware profiling}

\maketitle

% ============================================================================
\section{Introduction}
% ============================================================================

GPU kernel optimization is one of the most expertise-intensive tasks in high-performance computing. Achieving near-peak performance on modern GPUs requires navigating a complex space of decisions: memory hierarchy exploitation, parallelism patterns, hardware-specific features (Tensor Cores, shared memory), and algorithm-level transformations (online algorithms, kernel fusion). This knowledge is typically \emph{implicit}---residing in the minds of expert programmers---and \emph{fragmented}---scattered across papers, library implementations, and tribal knowledge.

Recent work has explored two approaches to automate this process. \emph{Rule-based systems}~\cite{} encode expert knowledge as static heuristics or search templates, but cannot generalize beyond their predefined rules. \emph{Search-based systems}~\cite{} explore the optimization space through autotuning or reinforcement learning, but treat each kernel independently, discarding the structural similarity between optimization strategies.

We make a key observation: \textbf{HPC kernel optimization has reusable structure}. The same optimization \emph{pattern}---such as ``tile over the reduction dimension, compute partial results via GEMM in shared memory, apply an online algorithm to avoid materializing intermediates''---applies to KMeans (distance + argmin), N-Body (force + accumulation), and GMM (likelihood + logsumexp), despite these being entirely different computational problems. We call such reusable patterns \emph{optimization skills}.

This observation leads to three claims:

\begin{enumerate}
    \item \textbf{Skill Discovery.} Optimization skills can be automatically discovered from optimization trajectories guided by hardware profiling, without manual engineering.
    \item \textbf{Skill Transfer.} Skills discovered on one kernel are transferable to unseen kernels that share similar computational structure.
    \item \textbf{Skill Compactness.} A small number of skills (5--10) can cover a large portion of the optimization space across the 7 Berkeley Dwarfs.
\end{enumerate}

To validate these claims, we build an LLM-driven agent that interacts with three interfaces:
\begin{itemize}
    \item \textbf{Knowledge} (K): retrieval, summarization, and accumulation of optimization patterns from papers, libraries, and prior experience.
    \item \textbf{Environment} (E): hardware profiling via NVIDIA Nsight Compute (NCU), providing bandwidth utilization, compute throughput, Tensor Core usage, and occupancy.
    \item \textbf{Objective} (O): multi-scale benchmarking with version tracking, regression detection, and library ceiling comparison.
\end{itemize}

We evaluate on 10 operators spanning all 7 Berkeley Dwarfs (dense/sparse linear algebra, spectral methods, N-body, structured/unstructured grids, MapReduce) on an NVIDIA A100 GPU. Our results show:

\begin{itemize}
    \item 5 skills cover 10 operators across 7 Dwarfs, achieving 1.18--6.04$\times$ speedup where applicable.
    \item The \texttt{tiled\_gemm\_online\_reduce} skill, discovered on KMeans, transfers directly to N-Body (2.19$\times$) and GMM (1.44$\times$).
    \item NCU-guided diagnosis enables the agent to progress from 0.07$\times$ (broken kernel) to 1.44$\times$ (correct kernel) in a single iteration, compared to blind trial-and-error.
    \item The agent correctly identifies 4 out of 5 cases where IO optimization is \emph{not} effective (Softmax, Conv2D, SpMV, KMeans-scalar), avoiding wasted effort.
\end{itemize}

% ============================================================================
\section{Background and Motivation}
% ============================================================================

\subsection{The IO Optimization Hypothesis}

Flash Attention~\cite{dao2022flashattention} demonstrated that reducing HBM memory traffic via kernel fusion and online algorithms can yield dramatic speedups for memory-bound GPU operations. This ``IO-aware'' approach has been extended to other operators~\cite{} with significant success. However, we find that \textbf{IO reduction is necessary but not sufficient}:

\begin{table}[h]
\centering
\caption{IO reduction vs.\ actual speedup on A100. IO optimization fails when the operator is compute-bound, data fits in L2 cache, or the optimization destroys GEMM parallelism.}
\label{tab:io-vs-speedup}
\begin{tabular}{lccl}
\toprule
Operator & IO $\downarrow$ & Speedup & Root Cause of Mismatch \\
\midrule
CrossEntropy & 80\% & \textbf{6.04$\times$} & --- (memory-bound, data $\gg$ L2) \\
LayerNorm & 50\% & \textbf{4.58$\times$} & --- (memory-bound) \\
Stencil2D & 99\% & \textbf{4.74$\times$} & --- (temporal tiling effective) \\
N-Body & 100\% & \textbf{2.19$\times$} & --- (SMEM tiling eliminates $N^2$) \\
KMeans (tl.dot) & 94\% & \textbf{1.44$\times$} & --- (GEMM structure preserved) \\
\midrule
KMeans (scalar) & 94\% & 0.07$\times$ & Lost Tensor Core (Dim 4) \\
Softmax & 65\% & 0.86$\times$ & Data fits L2 cache (Dim 3) \\
Conv2D & 86\% & 0.57$\times$ & cuDNN already optimal (Dim 6) \\
SpMV & 0\% & 0.76$\times$ & Wrong pattern for data (K mismatch) \\
\bottomrule
\end{tabular}
\end{table}

This motivates a multi-dimensional analysis beyond IO complexity alone.

\subsection{The Skill Hypothesis}

We observe that successful optimizations share common \emph{skills} across different operators:

\begin{table}[h]
\centering
\caption{5 optimization skills cover 10 operators across 7 Berkeley Dwarfs.}
\label{tab:skills}
\begin{tabular}{lll}
\toprule
Skill & Operators & Dwarfs \\
\midrule
Online streaming reduce & CE, LN, GMM & MapReduce \\
Tiled GEMM + online reduce & KMeans, N-Body & Dense LA, N-Body \\
Temporal tiling & Stencil2D & Structured Grid \\
SMEM butterfly fusion & FFT & Spectral \\
Fused scatter-gather & Graph Laplacian & Unstructured Grid \\
\bottomrule
\end{tabular}
\end{table}

This compactness suggests that learning a small skill library could dramatically reduce the search space for new kernels.

% ============================================================================
\section{The KEO Framework}
% ============================================================================

\subsection{Formalization}

We formalize kernel optimization as an iterative process over three interfaces:

\textbf{Knowledge ($\mathcal{K}$).} A growing library of optimization skills $\mathcal{S} = \{s_1, \ldots, s_k\}$ where each skill $s = (c, p, t)$ consists of:
\begin{itemize}
    \item \emph{Condition} $c$: when the skill applies (e.g., ``operator is memory-bound with materialized $N \times K$ intermediate'')
    \item \emph{Pattern} $p$: the optimization structure (e.g., ``tile over $K$, compute partial GEMM via \texttt{tl.dot}, apply online argmin'')
    \item \emph{Transformation} $t$: the code transformation (e.g., a Triton kernel template)
\end{itemize}

\textbf{Environment ($\mathcal{E}$).} Hardware profiling metrics $e = (\text{BW}_\%, \text{Compute}_\%, \text{TC}, \text{Occ}_\%, \ldots)$ obtained from NVIDIA Nsight Compute (NCU), providing ground-truth utilization data.

\textbf{Objective ($\mathcal{O}$).} Performance evaluation $o = (\text{speedup}, \text{correctness}, \text{version\_history})$ with multi-scale benchmarking and library ceiling comparison.

The agent loop is:
\begin{equation}
    a_t = \pi(s_{1:t-1}, e_{1:t-1}, o_{1:t-1}; \mathcal{K})
\end{equation}
where $\pi$ is the LLM policy, $a_t$ is the action at step $t$ (analyze, fuse, generate kernel, profile, etc.), and $\mathcal{K}$ is updated after successful optimizations.

\subsection{Skill Lifecycle}

\begin{enumerate}
    \item \textbf{Discovery.} From an optimization trajectory $(a_1, e_1, o_1, \ldots, a_T, e_T, o_T)$ that achieves speedup $> 1\times$, extract the key (condition, pattern, transformation) triple.
    \item \textbf{Abstraction.} Generalize the skill by replacing operator-specific details with structural features (e.g., ``$N \times K$ intermediate'' $\to$ ``materialized pairwise computation'').
    \item \textbf{Transfer.} Match new operators against skill conditions; apply the highest-scoring skill.
    \item \textbf{Evolution.} Update skill success statistics; refine conditions based on failure cases (e.g., ``only when data $>$ L2 cache'').
\end{enumerate}

% ============================================================================
\section{System Design}
% ============================================================================

\subsection{Agent Architecture}

The agent exposes 20 tools organized by KEO interface:

\begin{itemize}
    \item \textbf{K tools:} \texttt{retrieve\_pattern}, \texttt{pre\_check} (7-dimension static analysis), \texttt{show\_actions}
    \item \textbf{E tools:} \texttt{ncu\_profile} (NCU hardware counters + torch.profiler), \texttt{library\_ceiling}, \texttt{compare\_profile}, \texttt{occupancy\_analysis}
    \item \textbf{O tools:} \texttt{benchmark\_kernel} (version tracking + regression rollback), \texttt{compile\_and\_test}, \texttt{autotune\_kernel}, \texttt{verify}
    \item \textbf{Graph tools:} \texttt{analyze}, \texttt{fuse\_ops}, \texttt{fuse\_and\_online}, \texttt{generate\_kernel} (Triton/CUDA codegen)
\end{itemize}

\subsection{NCU-Driven Diagnosis}

Hardware profiling provides ground-truth signals that static analysis cannot:

\begin{table}[h]
\centering
\caption{NCU metrics enable precise diagnosis of optimization failures.}
\label{tab:ncu}
\begin{tabular}{lcccl}
\toprule
Operator & DRAM\% & SM\% & Warps\% & Diagnosis \\
\midrule
CrossEntropy & 77 & 68 & 66 & Memory-bound, near optimal \\
LayerNorm & 80 & 47 & 95 & Memory-bound, near optimal \\
KMeans (scalar) & 0 & 75 & 99 & Scalar loop (no DRAM access!) \\
KMeans (tl.dot) & 1 & 15 & 35 & Compute-bound with TC \\
Softmax & 7 & 49 & 82 & Data in L2, IO analysis invalid \\
\bottomrule
\end{tabular}
\end{table}

The signal \texttt{warps=99\%, DRAM=0\%} for KMeans-scalar is \emph{impossible to detect without runtime profiling}---it reveals that the SM is 99\% busy doing scalar arithmetic (not memory accesses), despite the kernel having 94\% IO reduction on paper.

\subsection{Version Tracking and Regression Detection}

LLM-generated kernels can regress (e.g., 5.83$\times$ $\to$ 3.02$\times$ on a rewrite attempt). We maintain:
\begin{itemize}
    \item A version history with speedup per version
    \item Automatic rollback when new speedup $< 0.95 \times$ best
    \item Convergence detection: \texttt{done} only after \texttt{ncu\_profile} + \texttt{library\_ceiling} + \texttt{autotune}
\end{itemize}

% ============================================================================
\section{Evaluation}
% ============================================================================

\subsection{Experimental Setup}

\textbf{Hardware.} NVIDIA A100 80GB PCIe (2039 GB/s HBM, 19.5 TFLOPS FP32, 312 TFLOPS BF16 TC).

\textbf{Operators.} 10 operators covering all 7 Berkeley Dwarfs: CrossEntropy, LayerNorm, Softmax, KMeans, FFT, Conv2D, Stencil2D, N-Body, SpMV, Graph Laplacian.

\textbf{Agent.} GPT-5.2 via Copilot proxy, max 25--30 steps per operator.

\textbf{Baselines.} (1) Materialized PyTorch baseline (explicit intermediates), (2) Library ceiling (cuDNN/cuFFT/cuBLAS/ATen), (3) Static IO analysis only (no profiling).

\subsection{Skill Transfer (Claim 2)}

The \texttt{tiled\_gemm\_online\_reduce} skill discovered on KMeans transfers to:

\begin{table}[h]
\centering
\caption{Cross-operator skill transfer results.}
\label{tab:transfer}
\begin{tabular}{llcc}
\toprule
Skill Source & Transfer Target & w/o Skill & w/ Skill \\
\midrule
KMeans & N-Body & 0.07$\times$ & \textbf{2.19$\times$} \\
KMeans & GMM E-step & 0.11$\times$ & \textbf{1.44$\times$} \\
CrossEntropy & LayerNorm & --- & \textbf{4.58$\times$} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Skill Coverage (Claim 3)}

\begin{table}[h]
\centering
\caption{5 skills cover 10 operators across 7 Dwarfs.}
\label{tab:coverage}
\begin{tabular}{lcccc}
\toprule
\# Skills & Operators Covered & Dwarfs Covered & Avg.\ Speedup \\
\midrule
1 & 3 (CE, LN, GMM) & 1 & 5.0$\times$ \\
2 & 5 (+KMeans, N-Body) & 3 & 3.8$\times$ \\
3 & 6 (+Stencil) & 4 & 3.9$\times$ \\
4 & 7 (+FFT) & 5 & 3.5$\times$ \\
5 & 8 (+Graph Lap.) & 6 & 3.2$\times$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{NCU Impact (Profiling-Guided vs.\ Blind)}

\begin{table}[h]
\centering
\caption{Impact of NCU-driven diagnosis on optimization outcome.}
\label{tab:ncu-impact}
\begin{tabular}{lcc}
\toprule
Operator & Blind (no NCU) & NCU-Guided \\
\midrule
KMeans & 0.07$\times$ (stuck) & \textbf{1.44$\times$} (1 iteration) \\
Softmax & 0.50$\times$ (wasted effort) & 0.86$\times$ (correctly identified as unfixable) \\
CrossEntropy & 5.83$\times$ (3 iterations) & 5.83$\times$ (confirmed in 1 iteration) \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================================
\section{Related Work}
% ============================================================================

\textbf{Kernel optimization with LLMs.} KernelBench~\cite{}, KernelSkill~\cite{}, and STARK~\cite{} use LLMs to generate or optimize GPU kernels, but rely on predefined knowledge bases or human-curated skills. Our work differs in that skills are \emph{discovered} from optimization trajectories rather than manually specified.

\textbf{Autotuning and search.} Ansor~\cite{zheng2020ansor}, Halide~\cite{}, and TVM~\cite{chen2018tvm} use search-based optimization but treat each kernel independently. Our skill abstraction enables cross-kernel transfer, reducing search cost.

\textbf{IO-aware optimization.} FlashAttention~\cite{dao2022flashattention} and its extensions demonstrate the power of IO-aware kernel design. We extend this to a general framework applicable across all 7 Berkeley Dwarfs, while identifying the conditions under which IO optimization fails.

\textbf{Hardware profiling for optimization.} NVIDIA Nsight Compute~\cite{} provides detailed kernel metrics. We integrate NCU into the agent loop for data-driven diagnosis, showing that runtime profiling signals (e.g., \texttt{DRAM=0\%, warps=99\%}) are essential for correct optimization decisions.

% ============================================================================
\section{Conclusion}
% ============================================================================

We presented a framework for automatically discovering, abstracting, and transferring GPU kernel optimization skills. Our key finding is that HPC optimization has compact, reusable structure: 5 skills cover 10 operators across all 7 Berkeley Dwarfs. By grounding the optimization agent in hardware profiling (NCU) and structuring its reasoning through the Knowledge--Environment--Objective framework, we achieve reliable speedups of 1.2--6.0$\times$ while correctly avoiding unprofitable optimizations. The skill-based approach transforms kernel optimization from a per-operator search problem into a growing knowledge system that improves with experience.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
